---
title: "Bone Health Assessment Protocol in Men with Prostate Cancer"
author: "Anthony Cunningham"
date: "March 13, 2019"
output: beamer_presentation
---

```{r setup, include=FALSE, echo=FALSE}
library(dplyr)
library(ggplot2)

Burroughs <- read.csv("C:/Users/Anthony/OneDrive/Documents/STAT 6220 Burroughs/STAT 6220 Burroughs_2017-18 Combined.csv", 
                      header = TRUE)
Burroughs$Year <- as.character(Burroughs$Year)
Burroughs$Year[Burroughs$Year == "2017"] <- "Oct-Dec 2017"
Burroughs$Year[Burroughs$Year == "2018"] <- "Oct-Dec 2018"
Burroughs$Year <- as.factor(Burroughs$Year)

within90_table <- count(Burroughs, Year, within90)
Labs <- count(Burroughs, Year, Complete.Labs)
```

## Background

- Men with Prostate Cancer are typically treated with ADT
- Major Side Effect: Lower Bone Densities
- ADT Patients' Recommended Evaluation Protocol
- Rikka's Clinic had not strictly adhered to this (Why?)

## Rikka's Treatment

- In response, Rikka implemented a streamlined referral process that, with one click, ordered all four lab tests and scheduled the patient with a bone health specialist.
- Patient Data Collected Oct.-Dec. of 2016-2018 
    - 2017: 69 Total Patients (25 New) 
    - 2018: 76 Total Patients (15 New)
- 2016 Data not used per recommendation
- Rikka collected relevant data only on New Patients

## Objectives

- Increase in proportion of patients referred to bone health specialist
- Increase proportion of patients who undergo the complete lab set
- Increase urologist knowledge and decrease perceived barriers to implementing proper bone health assessment protocols

## Objective 1: Increase Patient Referrals

- Problem: Rikka does not have data about ADT Patients who have not been referred; cannot compare proportions of referred vs not referred
- Subset: New Patients (Patients' Initial Meeting with Rikka)
    - 2017: 25 New Patients
    - 2018: 15 New Patients
- Compare proportion of New Patients Seen within 90 Days of Initial ADT treatment

## Objective 1 (Cont.)

```{r, within90, echo=FALSE}
within90 <- rbind(c(8, 7), c(9, 16))
within90_cool <- rbind(c(8, 9, 15), c(9, 16, 25), c(15, 25, 40))
colnames(within90_cool) <- c("Post", "Pre", "Total")
rownames(within90_cool) <- c("Within 90", "Not Within 90", "Total")
within90_cool
fisher.test(within90)
```

## Objective 1 (Cont.)

```{r, within90 analysis, echo=FALSE}
## Stacked Bar Chart of Proportion of Patients Seen by NP Within 90 Days of Initial ADT Treatment
stack_within90 <- ggplot(within90_table) + geom_bar(aes(y = n, x = Year, fill = within90), stat = "identity", position = "fill") + 
  ylab("Percent  ") + 
  ggtitle("Proportion of New Patients Seen Within 90 Days \n of Initial ADT Treatment, By Year") + 
  labs(tag = "Figure 2.\n") + 
  theme(plot.title = element_text(size = 14, hjust = 0.5), 
        axis.title.y = element_text(angle = 0, vjust = 0.54), 
        plot.tag = element_text(size = 12), 
        plot.tag.position = "top", 
        text = element_text(size = 12)) +
  scale_fill_manual(values = c("#ccb85f","#000000")) + expand_limits(y = 1.05) + 
  annotate("text", x = 1, y = 1.05, label = "Proportion of 'Yes': 9/25") + 
  annotate("text", x = 2, y = 1.05, label = "Proportion of 'Yes': 8/15")
stack_within90
```

## Objective 1 (Cont.)

```{r, within90 Donut, echo=FALSE}
just_Year <- plyr::revalue(within90_table$Year, c("Oct-Dec 2017"="2017", "Oct-Dec 2018"="2018"))
## Donut Plot
donut_within90 <- ggplot(within90_table) + geom_bar(aes(y = n, x = 1, fill = within90), stat = "identity", position = "fill") + coord_polar(theta = "y") + 
  facet_wrap(~ Year) + theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), 
                             panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
                             panel.border = element_blank(), strip.background = element_blank(), 
                             strip.text = element_blank()) + xlim(0, 1.5) + 
  geom_text(aes(x = 0, y = 0, label = as.factor(just_Year)), size = 5, fontface = "bold") + 
  labs(title = "Proportion of New Patients Seen Within 90 Days \n of Initial ADT Treatment, By Year", 
       subtitle = "Note: Patient Data Collected October-December Within Each Year", 
       tag = "Figure 2.\n") +
  theme(plot.title = element_text(size = 14, hjust = 0.5), 
        plot.subtitle = element_text(size = 12, hjust=0.5), 
        plot.tag = element_text(size = 12), 
        plot.tag.position = "top", 
        legend.position = "bottom", 
        text = element_text(size = 12)) +
  scale_fill_manual(values = c("#ccb85f","#000000"))
donut_within90
```

## Objective 2: Increase Proportion of Complete Lab Sets

```{r, Labs analysis, echo=FALSE}
labs <- rbind(c(11, 4), c(4, 21))
labs_cool <- rbind(c(11, 4, 15), c(4, 21, 25), c(15, 25, 40))
colnames(labs_cool) <- c("Post", "Pre", "Total")
rownames(labs_cool) <- c("Complete Set", "Incomplete Set", "Total")
labs_cool
fisher.test(labs)
```

## Objective 2 (Cont.)

```{r, Labs Stack, echo=FALSE}
stack_lab <- ggplot(Labs) + geom_bar(aes(y = n, x = Year, fill = Complete.Labs), stat = "identity", position = "fill") + 
    labs(title = "Proportion of Complete vs Incomplete Lab Sets \n Among New Patients, By Year", 
         y = "Percent    ",
         fill = "Complete Labs", 
         tag = "Figure 1.\n") +
    theme(plot.title = element_text(size = 14, hjust = 0.5), 
          axis.title.y = element_text(angle = 0, vjust = 0.54), 
          plot.tag = element_text(size = 12), 
          plot.tag.position = "top", 
          text = element_text(size = 12)) +
    scale_fill_manual(values = c("#ccb85f","#000000")) + expand_limits(y = 1.05) + 
    annotate("text", x = 1, y = 1.05, label = "Proportion of 'Yes': 4/25") + 
    annotate("text", x = 2, y = 1.05, label = "Proportion of 'Yes': 11/15")
stack_lab
```

## Objective 2 (Cont.)

```{r, Labs Donut, echo=FALSE}
just_Year <- plyr::revalue(Labs$Year, c("Oct-Dec 2017"="2017", "Oct-Dec 2018"="2018"))
## Donut Plot
donut_lab <- ggplot(Labs) + geom_bar(aes(y = n, x = 1, fill = Complete.Labs), stat = "identity", position = "fill") + coord_polar(theta = "y") + 
  facet_wrap(~ Year) + theme(axis.title = element_blank(), axis.text = element_blank(), axis.ticks = element_blank(), 
                             panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
                             panel.border = element_blank(), strip.background = element_blank(), 
                             strip.text = element_blank()) + xlim(0, 1.5) + 
  geom_text(aes(x = 0, y = 0, label = as.factor(just_Year)), size = 5, fontface = "bold") + 
  labs(title = "Complete vs Incomplete Lab Sets: \n Proportion of New Patients, By Year", 
       subtitle = "Note: Patient Data Collected October-December Within Each Year", 
       fill = "Complete Labs",
       tag = "Figure 1.\n") +
  theme(plot.title = element_text(size = 14, hjust = 0.5), 
        plot.subtitle = element_text(size = 12, hjust = 0.5), 
        plot.tag = element_text(size = 12), 
        plot.tag.position = "top", 
        legend.position = "bottom", 
        text = element_text(size = 12)) +
  scale_fill_manual(values = c("#ccb85f","#000000"))
donut_lab
```
